### 2020-06-29 manuscript/internship_report.Rmd
last fixes

keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
wordcount         : "12808"

@@

One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]). Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM. Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR). The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM). Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named Bayesian microsaccade detection [BMD]). Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).  

@@

First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals. Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification.  

@@

knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 200)

@@

# Operationalization
The proposed algorithm aims to improve the drawbacks of previous HMM algorithms. Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data. Besides implementing the algorithm in the statistical programming language R @RCoreTeam2020, I will develop its concrete technical design exploratorily.  
The first step will be choosing an appropriate metric that allows separating the gaze data into events. @Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events. The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012]. The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.  
The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data. The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters. The output of the model will be probabilities for hidden states. The states can be interpreted as event labels for each sample. For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade. These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability). Moreover, the model will output the parameters of the distribution of labeled samples for each event. It will allow generating samples from these distributions to simulate gaze data.  
Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model. The parameters of the model will be varied across sensible ranges. Then, the model will be applied to the data. Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.  
The first hypothesis will be tested by comparing different generative models on benchmark data. The models will differ in their assumed number of events in the data. The model comparison will result in a preferred model which scores best on the evaluation criterion. The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.  
To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data. The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).

@@

## Eye Movement Metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018]. This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics. Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades (e.g., @Mihali2017), but as proposed earlier, they are not in the scope of gazeHMM.

@@

## The Generative Model
The generative model underlying gazeHMM is a multivariate hidden Markov model (HMM). It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits. Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.  
In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model. The initial state model contains probabilites for the first state of the hidden sequence. The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM. The response model encompasses distributions describing the response variables for every state in the model.  
In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal. The response variables are treated as conditionally independent on the states. Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was merely taken to keep the HMM simple and identifiable.  
Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation). Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions. Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution. Thus, the uniform distribution should distinguish fixations from the other events.  
The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, no time-varying covariates are included. Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.

@@

(ref:sim-ranges) Approximate Ranges of Response Variables Used to Generate Parameter Values in the Simulation
(ref:sim-ranges-note) Units are deg/s (velocity), deg/s^2^ (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value. Velocity ranges are based on event velocities reported in @Larsson2013. Since I would also like to test the algorithm on extreme data distributions, I extended the ranges beyond those found in typical eye movement data.

@@

  p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
    geom_point(position = position_dodge(0.25)) + facet_grid(cols = vars(state.est)) +
    scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 3, 4, 5)) +
    geom_hline(yintercept = 0.1, linetype = "dashed") +
    geom_hline(yintercept = 0.5, linetype = "dashed")

@@

print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 10)))

@@

print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 6)))

@@

### Variation of Starting Values
The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM. For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)), suggesting good recovery. Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure \@ref(fig:plot-acc-3)). For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)), which means they were moderately well recovered. As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)). Regarding the four state-HMM, RMdSPS of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)). The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)). Taken together, increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.

@@

print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 10)))

@@

print(plots.acc.234[[2]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 6)))

@@

(ref:plot-acc-3-expl) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three after state labels were switched post-hoc (exploratory analysis). Colours indicate the variation in starting values used to estimate parameters. Solid vertical lines symbolize medians and hinges the first and third quartile. Whiskers range from hinges to lowest/highest value within 1.5 times the IQR. Crosses represent outliers.

@@

print(plots.acc.expl[[1]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 10)))

@@

print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 6)))

@@

(ref:tab-rmsd-3) Event Duration Descriptives and RMSD between Algorithms and Human Coders for the Image Condition
(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.

@@

## Future Directions
Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009]. Potential problem of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters). Another option to model the correlation could be to include one of the response variables as a covariate of the other.  
Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred. This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.  
Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged. Then, the maximum averaged probability could be used to classify the samples into events. This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020]. However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.  
Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data. As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements. As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades. Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].  
Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events. This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006]. Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs. This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades. Drawback of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].  
Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM. This could potentially be achieved by using different response distributions or parameter optimization methods. Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.

>>>>

keywords          : "eye movements; eye-tracking; parameter recovery; dependent mixture models"
wordcount         : "12962"

@@

One instance is described in @Salvucci2000 and combines the HMM with a threshold approach (named "Identification by HMM" [I-HMM]). Samples are first labeled as fixations or saccades, depending on whether their velocity exceeds a threshold, and then reclassified by the HMM. Recently, @Pekkanen2017 developed an algorithm that filters the position of gaze samples through naive segmented linear regression (NSLR). The algorithm uses an HMM to parse the resulting segments into fixations, saccades, smooth pursuits, and PSOs based on their velocity and angle (named NSLR-HMM). Another version by @Mihali2017 uses a Bayesian HMM to separate microsaccades (short saccades during fixations) from motor noise based on sample velocity (named "Bayesian Microsaccade Detection" [BMD]). Moreover, @Houpt2018 developed a hierarchical approach that describes sample velocity and acceleration through an autoregression (AR) model, computes the regression weights through an HMM, and estimates the number of events with a beta-process (BP) from the data (named BP-AR-HMM).  

@@

First, I-HMM relies on setting an appropriate threshold, which can distort the results [@Blignaut2009; @Komogortsev2010; @Shic2008]. Second, the current implementation of NSLR-HMM requires human-coded data, which limits its applicability. It also inherits fixed parameters that prevent the algorithm to adapt to the individual- or task-specific signals. Third, BMD limits the classification to microsaccades which are irrelevant in many applications and sometimes even considered as noise [@Duchowski2017]. The opposite problem was observed for BP-AR-HMM: It tends to estimate an unreasonable number of events from the data of which many are considered as noise events. Therefore, the authors suggest using it as an exploratory tool followed by further event classification [@Houpt2018].  

@@

knitr::include_graphics(here("manuscript/internship_report_files/external-files/hmm-structure.png"), auto_pdf = T, dpi = 400)

@@

# Operationalization
The proposed algorithm aims to improve the drawbacks of previous HMM algorithms. Therefore, its classification method is oriented to four major goals: (a) it does not require parameter settings through the user (e.g., thresholds) or (b) human-labeled data as input; (c) it covers the most relevant eye movement events, namely fixations, saccades, smooth pursuit, and PSOs; (d) it confirms rather than explores the presence of events in the data. Besides implementing the algorithm in the statistical programming language R [@RCoreTeam2020], I will develop its concrete technical design exploratorily.  
The first step will be choosing an appropriate metric that allows separating the gaze data into events. @Zemblys2018 provide an overview of different gaze data metrics and compare how well they predict human-labeled events. The next step will choosing a filtering/smoothing method to reduce the noise in the gaze data [for a comparison of different gaze data filters, see @Spakov2012]. The criteria for choosing a metric and a filter will be its accuracy and not relying on user parameter settings.  
The generative part of the algorithm will allow users to build an HMM which matches their theoretical assumptions about the gaze data. The algorithm will take a preprocessed gaze data metric (e.g., filtered velocity) as input and optimize the model parameters. The output of the model will be probabilities for hidden states. The states can be interpreted as event labels for each sample. For instance, the model could assign to a sample a probability of 0.6 for belonging to a fixation and 0.4 for belonging to a saccade. These probabilities can be used to compute the hidden state sequence (e.g., by choosing events with the highest probability). Moreover, the model will output the parameters of the distribution of labeled samples for each event. It will allow generating samples from these distributions to simulate gaze data.  
Before testing the hypotheses, I will conduct a simulation study to estimate the parameter recovery of the generative model: Artificial eye movement data will be generated by the model. The parameters of the model will be varied across sensible ranges. Then, the model will be applied to the data. Finally, the estimated results of the model will be compared to the true parameters that have been used to generate the data.  
The first hypothesis will be tested by comparing different generative models on benchmark data. The models will differ in their assumed number of events in the data. The model comparison will result in a preferred model which scores best on the evaluation criterion. The preferred model is expected to match the theoretically assumed number of present events in the benchmark data.  
To test the second hypothesis, the proposed algorithm will be compared against other algorithms that have also been applied to benchmark data. The performance of the proposed algorithm will be assessed by metrics that are commonly used in classification (e.g., Cohen's Kappa).

@@

## Eye Movement Metrics
Many different metrics can be used to describe gaze data and eye movement events [@Zemblys2018]. Here, my goal was to find those metrics that separate the gaze data samples belonging to different events the best. However, many metrics rely on thresholds or window ranges that have to be set by the user [e.g., the distance between the mean position in a 100 ms window before and after each sample, see @Olsson2007; @Zemblys2018]. This can be problematic because such parameters are often set without theoretical justification and they differ substantially between metrics. Because they belong to the most basic metrics which do not require parameter settings, I used velocity, acceleration, and sample-to-sample angle [synonymous to relative or change in angle; @Larsson2013] in gazeHMM.  
Theoretically, these three metrics should separate eye movement events clearly. Fixations typically inherit samples with low velocity and acceleration [@Larsson2013]. Due to tremor, the angle between samples should not follow any direction but a random walk [@Duchowski2017]. In contrast, saccade samples usually have a high velocity and acceleration and follow the same direction. PSO samples tend to have moderate velocity and high acceleration since they occur between saccades and low velocity events [@Larsson2013]. They can be specifically distinguished by their change in direction clustered around 180 degrees [@Pekkanen2017]. Importantly, the oscillations depend on the resolution of the gaze recording: Eye-trackers with higher sampling frequency yield more changes in direction and more samples in between those changes. Those samples in between typically follow the same direction. Thus, with high sampling frequencies, PSO samples might also cluster around a sample-to-sample angle of zero with outliers around 180 degrees. Lastly, smooth pursuit samples have a moderate velocity but low acceleration (due to the smoothness) and like saccades they follow the same direction [@Larsson2013; @Leigh2015]. Other algorithms focus exclusively on classifying microsaccades [e.g., @Mihali2017], but as proposed earlier, they are not in the scope of gazeHMM.

@@

## The Generative Model
The generative model underlying gazeHMM is a multivariate HMM. It can contain between two and four states that correspond to different eye movement events: The first state always represents fixations, the second saccades, the third PSOs, and the fourth smooth pursuits. Thus, users can choose whether they would like to classify only fixations and saccades, or additionally PSOs and/or smooth pursuits.  
In general, HMMs consist of three submodels: An initial state model, a transition model, and a response model. The initial state model contains probabilites for the first state of the hidden sequence. The evolution of the sequence is in turn described by the trantision model, which comprises the probabilities for transitioning between different states in the HMM. The response model encompasses distributions describing the response variables for every state in the model.  
In gazeHMM, the response model has three response variables which are the velocity and acceleration signals obtained by the SG filter and the change in angle signal. The response variables are treated as conditionally independent on the states. Conditional independence might not accurately resemble the relationship between velocity and acceleration (which are naturally correlated). This step was merely taken to keep the HMM simple and identifiable.  
Previous algorithms using HMMs have used Gaussian distributions to describe velocity and acceleration signals (sometimes after log-transforming them). However, several reasons speak against choosing the Gaussian: First, both signals are usually positive (depending on the computation). Second, the distributions of both signals appear to be positively skewed conditionally on the states and third, to have variances increasing with their mean. Thus, instead of using the Gaussian, it could be more appropriate to describe velocity and acceleration with a distribution that follows these three properties. In gazeHMM, I use gamma distributions with a shape and scale parametrization for this purpose. It has to be noted that the gamma was chosen out of convenience and the best fitting distribution might be different between eye-trackers, subjects, and tasks.   
To model the sample-to-sample angle, I pursued a novel approach in gazeHMM: Using a mixture of von Mises distributions (with a mean and concentration parameter) and a uniform distribution. Both the distributions and the metric operate on the full unit circle (i.e., between 0 and $2\pi$), which should lead to rather symmetric distributions. Because the fixations change their direction similar to a random walk, their sample-to-sample angle can be modeled by a uniform distribution. Thus, the uniform distribution should distinguish fixations from the other events.  
The initial state model and the transition model use multinomial distributions (with a category for each state). For all three submodels of the HMM, no time-varying covariates are included. Since the implementation of gazeHMM allows for regression models for each submodel and state, this implies that only intercepts for each parameter are estimated.

@@

(ref:sim-ranges) Approximate Ranges of Response Variables Used to Generate Parameter Values in the Simulation
(ref:sim-ranges-note) Units are deg/s (velocity), deg/s (acceleration), and radians (angle). ~ indicates that the distribution has a peak at this value. Velocity ranges are based on event velocities reported in @Larsson2013. Since I would also like to test the algorithm on extreme data distributions, I extended the ranges beyond those found in typical eye movement data.

@@

  p <- ggplot(data = data.long, aes(x = par.est, y = RMdSPD, color = cond)) +
    geom_point(position = position_dodge(0.25)) + facet_wrap(vars(state.est)) +
    scale_x_discrete(name = "Estimated parameter", labels = names.pars.est) +
    scale_y_continuous(breaks = c(0.1, 0.5, 1, 1.5, 2, 2.5, 3, 4, 5)) +
    geom_hline(yintercept = 0.1, linetype = "dashed") +
    geom_hline(yintercept = 0.5, linetype = "dashed")

@@

print(plots.rmsd.234[[1]][[1]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[1]][[2]] + theme(axis.text = element_text(size = 6)))

@@

print(plots.rmsd.234[[1]][[3]] + theme(axis.text = element_text(size = 8)))

@@

### Variation of Starting Values
The third part of the simulation investigated how increasing the variation in generating starting values affected parameter recover and accuracy of the HMM. For the two-state HMM, all parameters displayed RMdSPDs lower than 0.1 (see Figure \@ref(fig:plot-rmdspd-3-2)), suggesting good recovery. Cohen's kappa values were slightly lower than 1, indicating almost perfect accuracy (see Figure \@ref(fig:plot-acc-3)). For the three-state HMM, RMdSPDs were lower than 0.1 except for $a_{12}$ and $a_{31}$, which were below 0.5 (see Figure \@ref(fig:plot-rmdspd-3-3)), which means they were moderately well recovered. As in previous parts, Cohen's kappa values were mostly above 0.95 (almost perfect accuracy) with exceptions clustering around zero and -0.33 (see Figure \@ref(fig:plot-acc-3)). Regarding the four state-HMM, RMdSPDs of transition probabilities for states one and four were clustered above 0.1, other estimated parameters in the model showed values below 0.1 (see Figure \@ref(fig:plot-rmdspd-3-4)). The HMM showed substantial to almost perfect accuracy, as Cohen's kappa values were mostly above 0.8 with a few values clustering around 0.6, 0.25, and zero (see Figure \@ref(fig:plot-acc-3)). Taken together, increasing the variation in starting values for parameter estimation did neither affect the parameter recovery nor the accuracy.

@@

print(plots.rmsd.234[[2]][[1]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.acc.234[[2]])

@@

print(plots.rmsd.234[[2]][[2]] + theme(axis.text = element_text(size = 6)))

@@

print(plots.rmsd.234[[2]][[3]] + theme(axis.text = element_text(size = 8)))

@@

(ref:plot-acc-3-expl) Boxplots displaying Cohen's kappa depending on the number of states in the HMM in part three after state labels were switched post-hoc (exploratory analysis). Colours indicate the variation in starting values used to estimate parameters. Solid lines symbolize medians and hinges the first and third quartile. Whiskers range from hinges to lowest/highest value within 1.5 times the IQR. Crosses represent outliers.

@@

print(plots.acc.expl[[1]])

@@

print(plots.rmsd.234[[3]][[1]] + theme(axis.text = element_text(size = 8)))

@@

print(plots.rmsd.234[[3]][[2]] + theme(axis.text = element_text(size = 6)))

@@

print(plots.rmsd.234[[3]][[3]] + theme(axis.text = element_text(size = 8)))

@@

(ref:tab-rmsd-3) Event Duration Descriptives and RMSD Between Algorithms and Human Coders for the Image Condition
(ref:tab-rmsd-3-note) Durations are displayed in seconds. gazeHMM classified three events. RMSD = root mean square deviation. Table adapted from @Andersson2017.

@@

## Future Directions
Despite its advantages, there are several aspects in which gazeHMM can be improved: First, a multivariate distribution could be used to account for the correlation between velocity and acceleration signals [for examples, see @Balakrishnan2009]. Potential problems of this approach might be choosing the right distribution and convergence issues (due to a large number of parameters). Another option to model the correlation could be to include one of the response variables as a covariate of the other.  
Second, instead of the gamma being the generic (and potentially inappropriate) response distribution, a non-parametric approach could be used: @Langrock2015 use a linear combination of standardized B-splines to approximate response densities, which led to HMMs with less states being preferred. This approach could potentially combat the problem of unexpectedly high-state HMMs being preferred for eye movement data.  
Third, one solution to diverging results when comparing gazeHMM with different events could be model averaging: Instead of using the maximum posterior state probability of each sample from the preferred model, the probabilities could be weighted according to a model selection criterion (e.g., Schwarz weight) and averaged. Then, the maximum averaged probability could be used to classify the samples into events. This approach could lead to a more robust classification because it reduces the overconfidence of each competing model and easily adapts to new data [analogous to Bayesian model averaging; @Hinne2020]. However, the model comparison for gazeHMM often showed extreme weights for a five-state model, which would lead to a very limited influence of the other models in the averaged probabilities.  
Fourth, including covariates of the transition probabilities and response parameters could improve the fit of gazeHMM on eye movement data. As pointed out earlier, just estimating intercepts of parameters could be too simple to model the complexity of eye movements. As candidate covariates I suggest periodic functions of time [@Li2017] which could, for instance, capture the specific pattern of saccades. Whether covariates are improving the fit of submodels to eye movement data could in turn be assessed by inspecting pseudo-residuals and auto-correlation functions [see @Zucchini2016].  
Fifth, to avoid rapid switching between fixations and smooth pursuits as well as unreasonably short saccades, gazeHMM could explicitly model the duration of events. This can be achieved by setting the diagonal transition probabilities to zero and assign a distribution of state durations to each state [@Bishop2006]. Consequently, the duration distributions of fixations and smooth pursuits could differ from saccades and PSOs. This extension of the HMM is also called hidden semi-Markov model and has been successfully used by @Mihali2017 to classify microsaccades. Drawback of this extension are higher computational costs and difficulties with including covariates [@Zucchini2016].  
Lastly, allowing constrained parameters in the HMM could replace some of the postprocessing steps in gazeHMM. This could potentially be achieved by using different response distributions or parameter optimization methods. Moreover, switching from the maximum likelihood to the Markov chain Monte Carlo (Bayesian) framework could help avoiding convergence problems with constrained parameters, but would also open new research questions about suitable priors for HMM parameters in the eye movement domain.
